---
title: "Bout Analysis and Sedentary Patterns"
author: "Paul R. Hibbing"
output: rmarkdown::pdf_document
urlcolor: blue
linkcolor: blue
vignette: >
  %\VignetteIndexEntry{Bout Analysis and Sedentary Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PBpatterns)
```

## Introduction and Installation

This vignette will show you how to use `PBpatterns` for analyzing bouts (of any
physical behavior) and sedentary patterns (specifically). The first step is
making sure you have the `PBpatterns` package installed on your computer. Here's
how:

```{r install, eval=FALSE}

## remotes is a package that makes it easy to install packages from GitHub, but
## in my experience it sometimes struggles to install the related packages (i.e.,
## dependencies) correctly. So first we'll do a manual workaround. All it's
## doing is looking through a list of required packages, and installing any of
## them that haven't already been installed (they'll be skipped if they have).
## Be aware: Some of these packages may have long installation times.

invisible(lapply(
  c(
    "DescTools", "ggplot2", "magrittr", "PAutilities", "purrr", "utils",
    "AGread", "PhysicalActivity", "tree", "randomForest", "knitr", "rmarkdown"
  ),
  function(x) if (!x %in% installed.packages()) install.packages(x)
))

## Once that's done, we can (hopefully) install from GitHub
remotes::install_github("paulhibbing/PBpatterns", dependencies = FALSE)

```

Copy and paste the above into your R console, then hit enter to run it.

## Preparation

After installation (and just like for the [CRIB
method](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf)),
all you need is some activity data and the `analyze_bouts` function. For this
demonstration, let's use some sample NHANES data.

```{r data, cache=TRUE}

data(example_data, package = "PBpatterns")

```

This dataset has activity counts that we can use to look at bouts of sedentary
behavior (SB), light physical activity (LPA), and moderate-to-vigorous physical
activity (MVPA). For illustration, let's say we initially coded our data as SB
($PAXINTEN \leq 100$), LPA ($PAXINTEN\:\:101-759$), or MVPA ($PAXINTEN \geq
760$).

\newpage

```{r prep, cache=TRUE}

## Determine minute-by-minute intensity
x <- cut(
  example_data$PAXINTEN,
  breaks = c(-Inf, 101, 760, Inf),
  labels = c("SB", "LPA", "MVPA"),
  right = FALSE
)

```

To see how we can use this in the `analyze_bouts` function, first it's a good
idea to view the help page for that function.

```{r help, eval=FALSE}

?PBpatterns::analyze_bouts

```

## Basic Usage and Available Methods

Any call to `analyze_bouts` will start with the same three elements: `x`,
`target`, and `method`.

```{r basics, eval=FALSE}

## This code is for illustration -- it will throw an informative error if you
## try to run it, but don't worry -- we will see how to use the rest of the code
## in a bit

PBpatterns::analyze_bouts(
  x = x,
  target = "MVPA",
  method = c(
    ## Choose from:
    "rle_standard",
    "CRIB",
    "Troiano_MVPA",
    "SB_summary",
    "MVPA_summary"
  )
)

```

The first couple of arguments are pretty straightforward:

* **x** is the data you want to analyze. It needs to be a factor variable. 
* **target** is the level of `x` for which you want the bout information.

The `method` argument is ever so slighltly more involved. As you can see, there
are currently five available methods:

* **rle_standard** This is the traditional method based on run-length encoding.
It simply returns information (start index, end index, and duration) about every
distinct occurrence.
* **CRIB** See the [CRIB vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf)
* **Troiano_MVPA** This is the method of [Troiano et al.
(2008)](https://journals.lww.com/acsm-msse/pages/articleviewer.aspx?year=2008&issue=01000&article=00025&type=Fulltext)
for assessing bouts of MVPA.
* **SB_summary** This is the option to select if you want to analyze sedentary
patterns. It's also the driving function behind the `profile_describe_sb`
described in the [sedentary profiles
vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/SBprofiles.pdf).
* **MVPA_summary** This method is similar to `SB_summary`, but simpler and
focused on MVPA instead of SB.

The first three methods return a data frame with one row per bout. The
`*_summary` methods return a one-row data frame that summarizes all of the
bouts.

## Completing the Call and Understanding the Output

Each method requires one more argument (`epoch_length_sec`) to run properly.
Additionally, you can provide values for arguments called `is_wear` (wear time
indicator), `valid_indices` (valid day indicator), and
`minimum_bout_duration_minutes` (the shortest allowable bout length) to them
all. Apart from that, there are specialized settings you can feed into each
method. This is where the help file (see `?PBpatterns::analyze_bouts`) is so
important, as noted above. In that file, you can see what the relevant arguments
are for each method. In most cases, there are well-defined default values, so
you probably won't need to provide any extra information. But it's still good to
know what's possible. In the case of `CRIB`, there are some arguments for which
a default value can't be defined. So you can expect to see informative errors if
you don't specify them all. Let's look at some code now.

### Run-Length Encoding Standard Method

```{r standard, cache=TRUE}

standard_bouts <- PBpatterns::analyze_bouts(
  x, "SB", "rle_standard", epoch_length_sec = 60
)

head(standard_bouts)

```

### CRIB

See the [CRIB vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf).

### Troiano MVPA

```{r troiano, cache=TRUE}

troiano_bouts <- PBpatterns::analyze_bouts(
  x, "MVPA", "Troiano_MVPA", epoch_length_sec = 60
)

head(troiano_bouts)

```

### SB Summary

```{r SB, cache=TRUE}

## Note the warning this gives about returning NA for the predicted usual bout
## duration

SB_patterns <- PBpatterns::analyze_bouts(
  x, "SB", "SB_summary", epoch_length_sec = 60
)

SB_patterns

```

For this method, the output yields many variables. Some are self-explanatory,
but others may be more cryptic (particularly those used for sedentary profiles).
Here are explanations for the ones that need it:

* The **Q\*\_bout** variables are bout length percentiles, in minutes.
* **IQR** and **IDR** are the interquartile and interdecile ranges,
respectively, in minutes
* **SB\_perc** is the percentage of total time that was spent sedentary
* **bout\_frequency** is given in bouts per hour of wear time
* **sb\_0\_14**, **sb\_15\_29**, and **sb\_30\_Inf** give total sedentary time
(minutes) in bouts of $< 15$ minutes, $15-29.9$ minutes, and $\geq 30$ minutes,
respectively
* **ubd\_empirical** is the usual bout duration (minutes), calculated from the
observed data
* **ubd\_predicted** is the usual bout duration (minutes), calculated using a
lonlinear modeling method
* **fragmentation\_index** is given as sedentary breaks per sedentary hour
* **gini** is the Gini index
* **alpha** is alpha from the power law distribution (see [Chastin & Granat
(2010)](https://www.sciencedirect.com/science/article/abs/pii/S096663620900602X))
* **alpha\_se** is the standard error for `alpha`

### MVPA Summary

```{r MVPA, cache=TRUE}

MVPA_patterns <- PBpatterns::analyze_bouts(
  x, "MVPA", "MVPA_summary", epoch_length_sec = 60
)

MVPA_patterns

```

## Further Tools for Sedentary Pattern Analysis

There are a couple more tools in `PBpatterns` that can be leveraged for research
focused on sedentary patterns. The first is `summarize_weartime`, and its
general usage looks like this:

```{r weartime, cache=TRUE}

## First, add a timestamp variable to example_data
example_data$timestamp <- seq(
  as.POSIXlt(Sys.Date()),
  by = "1 min",
  length.out = nrow(example_data)
)

## Then, add a random wear time indicator to `example_data` (in real life, you
## might use the `PhysicalActivity` package for this)
set.seed(610)
example_data$is_wear <- sample(c(FALSE, TRUE), nrow(example_data), TRUE)

## Now, run the function
PBpatterns::summarize_weartime(example_data, "is_wear", "timestamp")

```

On its own, this function is somewhat unremarkable. The real power comes into
play when we combine it with other package code. For the next chunk, we will
use the `purrr` package to apply functions separately for each day in the
`example_data` object, then combine the results. This is a concise approach, but
might be tough to follow -- Don't worry too much about the specifics. This is
just for illustration, and in the real world you can accomplish the same thing
using a [for
loop](https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r?utm_source=adwords_ppc&utm_medium=cpc&utm_campaignid=1658343521&utm_adgroupid=63833879975&utm_device=c&utm_keyword=for%20loop%20r&utm_matchtype=p&utm_network=g&utm_adpostion=&utm_creative=469789579350&utm_targetid=aud-517318241987:kwd-300177784001&utm_loc_interest_ms=&utm_loc_physical_ms=9023228&gclid=Cj0KCQiAjJOQBhCkARIsAEKMtO0WwT-kkNOrl9igPUceKrAjKurpqvUs4oTNFE56Lg5NuPwiBly1w-MaAqoyEALw_wcB)
or any other approach you're comfortable with. (Be aware of some [looping
limitations and
alternatives](https://www.r-bloggers.com/2016/01/strategies-to-speedup-r-code/),
though. In R, I prefer to use loops for saving an output data file in each
iteration, rather than appending the iteration's result to an existing object.)

```{r demo, cache=TRUE}

## Save intensity as a variable in the dataset
example_data$intensity <- x

## For simplicity, label each row of data as valid
example_data$valid_index <- TRUE

## Extract information about wear time, SB patterns, and MVPA (a common covariate)
weartime_info <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
  ~ summarize_weartime(.x, "is_wear", "timestamp", .x$valid_index)
)

sb_bouts <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
   ~ analyze_bouts(
     .x$intensity, "SB", "SB_summary",
     is_wear = .x$is_wear,
     valid_indices = .x$valid_index,
     epoch_length_sec = 60
    )
)

mvpa_bouts <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
  ~ analyze_bouts(
      .x$intensity, "MVPA", "MVPA_summary",
      is_wear = .x$is_wear,
      valid_indices = .x$valid_index,
      epoch_length_sec = 60
    )
)

## Now combine all the above pieces of information (This works because all the
## objects have matching and unique `epoch_length` and `total_weartime_min`
## columns). In real life, you wouldn't have a guarantee of this. Thus, you
## would need to set up the merge using additional indicators.

d <- merge(weartime_info, sb_bouts)
d <- merge(d, mvpa_bouts)

```

Now that we have our combined weartime/SB/MVPA dataset (the object called `d`),
we can use the `adjust_bout_summaries` function to calculate residualized
variables suitable for modeling.

```{r residualize, cache=TRUE}

## Set `verbose` to TRUE if you want console updates about what's happening
adjust_bout_summaries(d, verbose = FALSE)

```
This code added several variables:

* **SB\_hr\_day** is daily SB time (hours/day)
* **MVPA\_min\_day** is daily MVPA time (minutes/day) -- it's equivalent to
`total_MVPA_min` because of the way we set up this illustration
* **adj\_total\_SB** is adjusted total SB (hours/day)
* **adj\_mean\_SB\_bout** is adjusted mean SB bout length (minutes)
* **adj\_sb\_0\_14**, **adj\_sb\_15\_29**, and **adj\_sb\_30\_Inf** are adjusted
SB time (minutes) in bouts of $< 15$ minutes, $15-29.9$ minutes, and $\geq 30$
minutes, respectively
* **adj\_median\_sb\_bout** is the adjusted median bout duration (minutes)
* **adj\_MVPA** is adjusted MVPA time (minutes)

## Wrapping Up

This should get you on your way to using `PBpatterns` for your analyses. As
always, feel free to [post an
issue](https://github.com/paulhibbing/PBpatterns/issues/new/choose) if something
can be improved. This is a big effort, and a definite work in progress, so
suggestions and tips are appreciated!
