---
title: "Bout Analysis and Sedentary Patterns"
author: "Paul R. Hibbing"
output: rmarkdown::pdf_document
urlcolor: blue
linkcolor: blue
vignette: >
  %\VignetteIndexEntry{Bout Analysis and Sedentary Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PBpatterns)
```

## Introduction and Installation

This vignette will show you how to use `PBpatterns` for analyzing bouts (of any
physical behavior) and sedentary patterns (specifically). The first step is
making sure you have the `PBpatterns` package installed on your computer. Here's
how:

```{r install, eval=FALSE}

## remotes is a package that makes it easy to install packages from GitHub, but
## in my experience it sometimes struggles to install the related packages (i.e.,
## dependencies) correctly. So first we'll do a manual workaround. All it's
## doing is looking through a list of required packages, and installing any of
## them that haven't already been installed (they'll be skipped if they have).
## Be aware: Some of these packages may have long installation times.

invisible(lapply(
  c(
    "DescTools", "ggplot2", "magrittr", "PAutilities", "purrr", "utils",
    "AGread", "PhysicalActivity", "tree", "randomForest", "knitr", "rmarkdown"
  ),
  function(x) if (!x %in% installed.packages()) install.packages(x)
))

## Once that's done, we can (hopefully) install from GitHub
remotes::install_github("paulhibbing/PBpatterns", dependencies = FALSE)

```

Copy and paste the above into your R console, then hit enter to run it.

## Preparation

After installation (and just like for the [CRIB
method](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf)),
all you need is some activity data and the `analyze_bouts` function. For this
demonstration, let's use some sample NHANES data. We'll also do some convenient
pre-processing based on wear time analysis from the `PhysicalActivity` package.

```{r data, cache=FALSE}

## Load the data
data(example_data, package = "PBpatterns")

## Check that `PhysicalActivity` is installed
if (!"PhysicalActivity" %in% installed.packages()) install.packages(
  "PhysicalActivity"
)

## Add a timestamp variable to satisfy `PhysicalActivity` requirements
example_data$TimeStamp <- seq(
  as.POSIXlt("2020-02-02", "UTC"),
  by = "1 min",
  length.out = nrow(example_data)
)

## Now label wear minutes
example_data <- PhysicalActivity::wearingMarking(
  example_data, perMinuteCts = 1, cts = "PAXINTEN",
  newcolname = "is_wear", getMinuteMarking = TRUE
)

## Format the `is_wear` variable as logical
example_data$is_wear <- example_data$is_wear == "w"

## And lastly, label valid indices (i.e., those that occur on days with at least
## 10 h of wear time). This is a bit of an abstract call -- you can think of it
## as a loop that labels each individual element of `PAXDAY` with the total wear
## time for that whole day.
example_data$valid_index <- sapply(
  example_data$PAXDAY,
  function(i, vec) vec[match(i, names(vec))],
  vec = tapply(example_data$is_wear, example_data$PAXDAY, sum)
) >= 600

## Here's a summary of what this tells us:
ftable(xtabs(~PAXDAY+valid_index+is_wear, example_data))

```

All the days are valid based on our criteria, with wear time ranging from just
over 10 hours (617 minutes on day 5) to around 14.5 hours (874 minutes on day 6).

Now, let's look at intensity classifications and subsequent analysis of
intensity bouts. This dataset has activity counts that we can use to look at
bouts of sedentary behavior (SB), light physical activity (LPA), and
moderate-to-vigorous physical activity (MVPA). For illustration, let's say we
initially coded our data as SB ($PAXINTEN \leq 100$), LPA
($PAXINTEN\:\:101-759$), or MVPA ($PAXINTEN \geq 760$).

```{r prep, cache=FALSE}

## Determine minute-by-minute intensity
example_data$intensity <- cut(
  example_data$PAXINTEN,
  breaks = c(-Inf, 101, 760, Inf),
  labels = c("SB", "LPA", "MVPA"),
  right = FALSE
)

```

\newpage

To see how we can use this in the `analyze_bouts` function, first it's a good
idea to view the help page for that function.

```{r help, eval=FALSE}

?PBpatterns::analyze_bouts

```

## Basic Usage and Available Methods

Any call to `analyze_bouts` will start with the same three elements: `x`,
`target`, and `method`.

```{r basics, eval=FALSE}

## This code is for illustration only. It will throw an informative error if you
## try to run it, but don't worry -- We will see how to use the rest of the code
## in a bit

PBpatterns::analyze_bouts(
  x = x,
  target = "MVPA",
  method = c(
    ## Choose from:
    "rle_standard",
    "CRIB",
    "Troiano_MVPA",
    "SB_summary",
    "MVPA_summary"
  )
)

```

The first couple of arguments are pretty straightforward:

* **x** is the data you want to analyze. It needs to be a factor variable. 
* **target** is the level of `x` for which you want the bout information.

The `method` argument is ever so slighltly more involved. As you can see, there
are currently five available methods:

* **rle_standard** This is the traditional method based on run-length encoding.
It simply returns information (start index, end index, and duration) about every
distinct occurrence.
* **CRIB** See the [CRIB vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf)
* **Troiano_MVPA** This is the method of [Troiano et al.
(2008)](https://journals.lww.com/acsm-msse/pages/articleviewer.aspx?year=2008&issue=01000&article=00025&type=Fulltext)
for assessing bouts of MVPA.
* **SB_summary** This is the option to select if you want to analyze sedentary
patterns. It's also the driving function behind the `profile_describe_sb`
described in the [sedentary profiles
vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/SBprofiles.pdf).
* **MVPA_summary** This method is similar to `SB_summary`, but simpler and
focused on MVPA instead of SB.

The first three methods return a data frame with one row per bout. The
`*_summary` methods return a one-row data frame that summarizes all of the
bouts.

## Completing the Call and Understanding the Output

Each method requires one more argument (`epoch_length_sec`) to run properly.
Additionally, you can provide values for arguments called `is_wear` (wear time
indicator), `valid_indices` (valid day indicator), and
`minimum_bout_duration_minutes` (the shortest allowable bout length) to them
all. Apart from that, there are specialized settings you can feed into each
method. This is where the help file (see `?PBpatterns::analyze_bouts`) is so
important, as noted above. In that file, you can see what the relevant arguments
are for each method. In most cases, there are well-defined default values, so
you probably won't need to provide any extra information. But it's still good to
know what's possible. In the case of `CRIB`, there are some arguments for which
a default value can't be defined. So you can expect to see informative errors if
you don't specify them all. Let's look at some code now.

### Run-Length Encoding Standard Method

```{r standard, cache=FALSE}

standard_bouts <- PBpatterns::analyze_bouts(
  example_data$intensity, "SB", "rle_standard", epoch_length_sec = 60
)

head(standard_bouts)

```

### CRIB

See the [CRIB vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/CRIB.pdf).

### Troiano MVPA

```{r troiano, cache=FALSE}

troiano_bouts <- PBpatterns::analyze_bouts(
  example_data$intensity, "MVPA", "Troiano_MVPA", epoch_length_sec = 60
)

head(troiano_bouts)

```

\newpage

### SB Summary {#SBsummary}

```{r SB, cache=FALSE}

## Note that here, we are providing extra information about wear time and valid
## days. This is optional, but it is a really good idea. In this case, if we
## don't provide this information, we will get a warning about failure to
## calculate the predicted usual bout duration.

SB_patterns <- PBpatterns::analyze_bouts(
  example_data$intensity, "SB", "SB_summary",
  is_wear = example_data$is_wear,
  valid_indices = example_data$valid_index,
  epoch_length_sec = 60
)

SB_patterns

```

For this method, the output yields many variables. Some are self-explanatory,
but others may be more cryptic (particularly those used for sedentary profiles).
Here are explanations for the ones that need it:

* **SB\_bout\_exclusion\_threshold\_minutes** reflects the setting that was
provided for \newline `minimum_bout_duration_minutes` -- It is renamed in the output to
avoid misinterpretation. (The latter term is clear when making the function
call, but not necessarily when interpreting the output)
* The **Q\*\_bout** variables are bout length percentiles, in minutes.
* **IQR** and **IDR** are the interquartile and interdecile ranges,
respectively, in minutes
* **SB\_perc** is the percentage of total time that was spent sedentary
* **bout\_frequency** is given in bouts per hour of wear time
* **sb\_0\_14**, **sb\_15\_29**, and **sb\_30\_Inf** give total sedentary time
(minutes) in bouts of $< 15$ minutes, $15-29.9$ minutes, and $\geq 30$ minutes,
respectively
* **ubd\_empirical** is the usual bout duration (minutes), calculated from the
observed data
* **ubd\_predicted** is the usual bout duration (minutes), calculated using a
lonlinear modeling method
* **fragmentation\_index** is given as sedentary breaks per sedentary hour
* **gini** is the Gini index
* **alpha** is alpha from the power law distribution (see [Chastin & Granat
(2010)](https://www.sciencedirect.com/science/article/abs/pii/S096663620900602X))
* **alpha\_se** is the standard error for `alpha`

\newpage

### MVPA Summary

```{r MVPA, cache=FALSE}

MVPA_patterns <- PBpatterns::analyze_bouts(
  example_data$intensity, "MVPA", "MVPA_summary", epoch_length_sec = 60
)

MVPA_patterns

```
Interpretation of this output is pretty straightforward. The
`MVPA_bout_exclusion_threshold_minutes` variable is interpreted the same way as
its counterpart discussed in [SB Summary](#SBsummary) above.

## Further Tools for Sedentary Pattern Analysis

There are a couple more tools in `PBpatterns` that can be leveraged for research
focused on sedentary patterns. The first is `summarize_weartime`, and its
general usage looks like this:

```{r weartime, cache=FALSE}

PBpatterns::summarize_weartime(example_data, "is_wear", "TimeStamp")

```

On its own, this function is somewhat unremarkable. The real power comes into
play when we combine it with other package code. For the next chunk, we will
use the `purrr` package to apply functions separately for each day in the
`example_data` object, then combine the results. This is a concise approach, but
might be tough to follow -- Don't worry too much about the specifics. This is
just for illustration, and in the real world you can accomplish the same thing
using a [for
loop](https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r?utm_source=adwords_ppc&utm_medium=cpc&utm_campaignid=1658343521&utm_adgroupid=63833879975&utm_device=c&utm_keyword=for%20loop%20r&utm_matchtype=p&utm_network=g&utm_adpostion=&utm_creative=469789579350&utm_targetid=aud-517318241987:kwd-300177784001&utm_loc_interest_ms=&utm_loc_physical_ms=9023228&gclid=Cj0KCQiAjJOQBhCkARIsAEKMtO0WwT-kkNOrl9igPUceKrAjKurpqvUs4oTNFE56Lg5NuPwiBly1w-MaAqoyEALw_wcB)
or any other approach you're comfortable with. (Be aware of some [looping
limitations and
alternatives](https://www.r-bloggers.com/2016/01/strategies-to-speedup-r-code/),
though. In R, I prefer to use loops for saving an output data file in each
iteration, rather than appending the iteration's result to an existing object.)

```{r demo, cache=FALSE}

## Extract information about wear time, SB patterns, and MVPA (a common
## covariate in sedentary pattern analysis)
weartime_info <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
  ~ summarize_weartime(.x, "is_wear", "TimeStamp", .x$valid_index)
)

sb_bouts <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
   ~ analyze_bouts(
     .x$intensity, "SB", "SB_summary",
     is_wear = .x$is_wear,
     valid_indices = .x$valid_index,
     epoch_length_sec = 60
    )
)


mvpa_bouts <- purrr::map_df(
  split(example_data, example_data$PAXDAY),
  ~ analyze_bouts(
      .x$intensity, "MVPA", "MVPA_summary",
      is_wear = .x$is_wear,
      valid_indices = .x$valid_index,
      epoch_length_sec = 60
    )
)

## Now combine all the above pieces of information (This works because all the
## objects have matching and unique `epoch_length` and `total_weartime_min`
## columns). In real life, you wouldn't have a guarantee of this. Thus, you
## would need to set up the merge using additional indicators (participant IDs
## and potentially dates as well).

d <- merge(weartime_info, sb_bouts)
d <- merge(d, mvpa_bouts)

```

Now that we have our combined weartime/SB/MVPA dataset (the object called `d`),
we can use the `adjust_bout_summaries` function to calculate residualized
variables suitable for modeling.

```{r residualize, cache=FALSE}

## Set `verbose` to TRUE if you want console updates about what's happening
adjust_bout_summaries(d, verbose = FALSE)

```
\newpage

The preceding code added several variables:

* **SB\_hr\_day** is daily SB time (hours/day)
* **sb\_0\_14\_hr\_day**, **sb\_15\_29\_hr\_day**, and **sb\_30\_Inf\_hr\_day**
are sedentary time (hours/day) in bouts of $< 15$ minutes, $15-29.9$ minutes,
and $\geq 30$ minutes, respectively
* **MVPA\_min\_day** is daily MVPA time (minutes/day) -- it's equivalent to
`total_MVPA_min` because of the way we set up this illustration
* **adj\_total\_SB** is adjusted total SB (hours/day)
* **adj\_mean\_SB\_bout** is adjusted mean SB bout length (minutes)
* **adj\_sb\_0\_14**, **adj\_sb\_15\_29**, and **adj\_sb\_30\_Inf** are adjusted
SB time (hours/day) in bouts of $< 15$ minutes, $15-29.9$ minutes, and $\geq 30$
minutes, respectively
* **adj\_median\_sb\_bout** is the adjusted median bout duration (minutes)
* **adj\_MVPA** is adjusted MVPA time (minutes)

## Wrapping Up

This should get you on your way to using `PBpatterns` for your analyses. As
always, feel free to [post an
issue](https://github.com/paulhibbing/PBpatterns/issues/new/choose) if something
can be improved. This definitely a work in progress, so suggestions and tips are
appreciated!
