---
title: "Clustered Recognition of Interrupted Bouts (CRIB)"
author: "Paul R. Hibbing"
output: rmarkdown::pdf_document
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{CRIB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PBpatterns)
```

## Introduction and Installation

This vignette will show you how to implement the CRIB method. The first step is
making sure you have the `PBpatterns` package installed on your computer. Here's
how:

```{r install, eval=FALSE}

## `remotes` is a package that makes it easy to install packages from GitHub, but
## in my experience it sometimes struggles to install the related packages (i.e.,
## dependencies) correctly. So first we'll do a manual workaround. All it's
## doing is looking through a list of required packages, and installing any of
## them that haven't already been installed (they'll be skipped if they have).
## Be aware: Some of these packages may have long installation times.

invisible(lapply(
  c(
    "DescTools", "ggplot2", "magrittr", "PAutilities", "purrr", "utils",
    "AGread", "PhysicalActivity", "tree", "randomForest", "knitr", "rmarkdown"
  ),
  function(x) if (!x %in% installed.packages()) install.packages(x)
))

## Once that's done, we can (hopefully) install from GitHub
remotes::install_github("paulhibbing/PBpatterns", dependencies = FALSE)

```

Copy and paste the above into your R console, then hit enter to run it.

## Preparation

Once you have the package installed, all you need is some activity data and the
`analyze_bouts` function. For this demonstration, let's use some sample NHANES
data.

```{r data, cache=FALSE}

data(example_data, package = "PBpatterns")

```

This dataset has activity counts that we can use to look at bouts of
moderate-to-vigorous physical activity (MVPA). For illustration, let's say we
initially coded our data as sedentary behavior ($PAXINTEN \leq 100$), light
physical activity ($PAXINTEN\:\:101-759$), or MVPA ($PAXINTEN \geq 760$).

\newpage

```{r prep, cache=FALSE}

## Determine minute-by-minute intensity
intensity <- cut(
  example_data$PAXINTEN,
  breaks = c(-Inf, 101, 760, Inf),
  labels = c("SB", "LPA", "MVPA"),
  right = FALSE
)

```

## Running the Code

Once we have our data (`intensity` in this case), we can plug
it into `analyze_bouts`. Let's see the code first, then go over what it means.

```{r bouts, cache=FALSE}

mvpa_bouts <- PBpatterns::analyze_bouts(
  x = intensity,
  target = "MVPA",
  method = "CRIB",
  target_buffer_mins = 10,
  longest_allowable_interruption_mins = 2,
  required_percent = 80,
  max_n_interruptions = Inf,
  minimum_bout_duration_minutes = 10,
  epoch_length_sec = 60
)

```

Here is what each piece means:

* **mvpa_bouts <-** Store the function results in an object called `mvpa_bouts`
* **PBpatterns::analyze_bouts** This tells R to find the `analyze_bouts` function
  in the `PBpatterns` package. In fact, if you run `PBpatterns::analyze_bouts`
  in your console, R will print the source code.
* **x = intensity** Here we specify that our input datastream is `intensity`,
  as defined in the earlier code.
* **target = "MVPA"** Here we specify which behavior we are interested in.
  The input data (`intensity`) has values in the set `{SB, LPA, MVPA}`, and
  we would like to look specifically at bouts of `MVPA`, with the other
  behaviors being lumped together in a single group called `other`.
* **method = "CRIB"** Here we specify that R should run the CRIB method. To
  explore the other options, check out the [Bout Analysis and Sedentary Patterns
  vignette](https://github.com/paulhibbing/PBpatterns/blob/main/vignettes/Bouts_and_Patterns.pdf).
* **target_buffer_mins = 10** Here we specify how our data should be
  stratified/partitioned. In this case, `intensity` will be
  stratified/partitioned anytime we see $\geq10$ consecutive minutes of `other`
  behavior.
* **longest_allowable_interruption_mins = 2** Here we specify that a valid bout
  should not include any single interruption lasting longer than 2 minutes.
* **required_percent = 80** Here we specify that a valid bout should be
  interrupted for no more than 20% of its full duration.
* **max_n_interruptions = Inf** Here we specify that a valid bout can have
  unlimited interruptions as long as the criteria for
  `longest_allowable_interruption_mins` and `required_percent` are met.
* **minimum_bout_duration_minutes = 10** Here we specify that only bouts lasting
  $\geq10$ min should be included in the output.
* **epoch_length_sec = 60** Here we specify what the epoch length of `intensity` is.
  
\newpage
  
The above elements are set up to allow flexible bout criteria depending on the
research question and the variable of interest. In our example, we set
`max_n_interruptions = Inf` to avoid a restriction in that area -- similar
approaches can be taken for other settings as well, by setting them to `0` or
`Inf` as appropriate. (For `required_percent`, 100 is the upper limit rather
than `Inf`.) Notably, `minimum_bout_duration_minutes` is a filtering criterion.
It has no direct effect on how the bouts are defined; it simply affects which
ones are retained *after* they have been defined.
  
## Interpreting the Output

Now let's take a look at the output and go over what it means:

```{r output, cache=FALSE}
mvpa_bouts
```

This is a data frame with one row per bout. The variables are:

* **start_index** The starting point of the bout (e.g., `intensity[912]` for
  the first bout in this example)
* **end_index** The ending point of the bout (e.g., `intensity[932]` for
  the first bout in this example)
* **values** A meaningless constant (equal to the setting of `target`),
  left over from run length encoding
* **n_total_events** The number of distinct behavior events occurring between
  `start_index` and `end_index`
* **n_value_events** The number of distinct target behavior events occurring
  between `start_index` and `end_index` (referred to as `value` events in
  reference to the `values` column)
* **n_interruption_events** The number of distinct interruption events
  occurring between `start_index` and `end_index`
* **overall_minutes** The combined duration of all `value`
  and `interruption` events
* **engaged_minutes** The combined duration of all `value` events
* **percent_time_engaged** Percentage of `length_total` comprised
  by `length_value`
* **total_interruption_minutes** The combined duration of
  all `interruption` events
* **longest_interruption_minutes** The duration of the single longest
  interruption event

\newpage

## Expanding the Output

In some cases we may want to convert our bout information back to the original
length of the input (i.e., `intensity`). We can use the `expand_bouts` function
to accomplish that.

```{r expansion, cache=FALSE}
expanded <- PBpatterns::expand_bouts(mvpa_bouts)
str(expanded)
table(expanded)
```

And we can also append that new variable into our original dataset as well.

```{r expansion2, cache=FALSE}
example_data$intensity <- expanded
head(example_data)
```

\newpage

## Plotting the output

If you want to visualize the results of your bout analysis, you can use a basic
`plot` function.

```{r plotting, cache=FALSE, fig.width=7, fig.height=5, fig.align='center'}
plot(mvpa_bouts)
```

## Conclusion

This should give you a broad sense of how to use the CRIB method and what else
you can do with it. Feel free to [post an issue on the GitHub
page](https://github.com/paulhibbing/PBpatterns/issues/new/choose) if any of the
above gives you trouble. Good luck!
